---
title: "Mutation_Signatures"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Mutation Signature Analysis
This rmarkdown describes the process to extract mutation signatures from sperm and identify mutagenic processes acting on bulk sperm across different age ranges. 

Mutation signatures will be extracted using the [*mutSignatures*](https://cran.r-project.org/web/packages/mutSignatures/vignettes/get_sarted_with_mutSignatures.html) package in R. 

Information on known COSMIC signatures (n=30) can be found [here](https://cancer.sanger.ac.uk/signatures/signatures_v2/)

## Load in packages
```{r, message=FALSE}
library(ggplot2)
library(mutSignatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(data.table)
```


## Generate mutationCounts object
```{r, message=FALSE}
## Define working directory
setwd("~/quinlan/u1240855/spermseq/scripts/analysis/mutation_signatures/")

## Import and prepare data from VCF files
file1 <- "19610X1.1.custom_variants_filtered.vcf"
file2 <- "19610X2.1.custom_variants_filtered.vcf"
files <- c(file1, file2)
vcf_data <- importVCFfiles(files)

## Load the UCSC GRCh38 reference genome  
grch38 <- BSgenome.Hsapiens.UCSC.hg38

## Compute mutation types
full_data <- processVCFdata(vcfData = vcf_data, BSGenomeDb = grch38, sample_colName = "SAMPLEID")

## Count mutation types in each sample
allVCF.counts <- countMutTypes(mutTable = full_data, sample_colName = "SAMPLEID")
```

## Examine *de novo* mutation signatures
```{r, results='asis'}
## Extract number of de novo signatures to extract
allVCF.assess <- prelimProcessAssess(input = allVCF.counts, approach = "counts")
parameter_num <- allVCF.assess[which(allVCF.assess$percentErr == 0)[1], "numProcess"]

## How many signatures should we extract? 
num_sign <- parameter_num

## Define parameters for the non-negative matrix factorization procedure
allVCF.params <- 
  setMutClusterParams(
    num_sign, 
    num_totIterations = 50, 
    num_parallelCores = 8
  )

## Extracct new signatures
allVCF.analysis <- decipherMutationalProcesses(input = allVCF.counts, 
                                               params = allVCF.params)

## Visualize exposures
deNovo.signs <- allVCF.analysis$Results$signatures
```
## *De Novo* Signatures {.tabset}
```{r, results='asis', echo=FALSE}
## Iterate through each signature and plot
for (num in 1:num_sign) {
  cat("###", paste0("De Novo Signature ", num), " \n")
  msigPlot(deNovo.signs, signature = num)
  cat(' \n\n')
}
```

## Examine known COSMIC signatures 
```{r}
## Retrieve COSMIC signatures from online repo
cosmix <- getCosmicSignatures()

## Breakdown signatures
DNA_damage <- c(1, 2, 7, 11, 13) ## 1 - spontaneous deamination, 2 - APOBEc cytosine deaminase, 7 - UV, 11 - alkylating agent, 13 - AID/APOBEC cytidine deaminase (C>U)
DNA_damage_df <- data.table("Signature" = DNA_damage, "Name" = rep("DNA Damage", length(DNA_damage)))

DNA_repair <- c(3, 6, 15, 20, 26) ## 3 - DNA DSB break repair, 6 - Mismatch repair, 15 - Mismatch repair, 20 - Mismatch repair, 26 - Mismatch repair
DNA_repair_df <- data.table("Signature" = DNA_repair, "Name" = rep("DNA Repair", length(DNA_repair)))

Replication <- c(9, 10) ## 9 - polymearse and AID, polymerase POLE
Replication_df <- data.table("Signature" = Replication, "Name" = rep("Replication", length(Replication)))

Lifestyle <- c(4, 22, 24, 29) ## 4 - smoking, 22 - aristolochic acid, 24 - aflatoxin, 29 - chewing tobacco
Lifestyle_df <- data.table("Signature" = Lifestyle, "Name" = rep("Lifestyle", length(Lifestyle)))

Unknown <- c(5, 8, 12, 14, 16, 17, 18, 19, 21, 23, 25, 27, 28, 30)
Unknown_df <- data.table("Signature" = Unknown, "Name" = rep("Unknown", length(Unknown)))

COSMIC_df <- rbind(DNA_damage_df, DNA_repair_df, Replication_df, Lifestyle_df, Unknown_df)
```

## COSMIC Signatures {.tabset}
```{r, results='asis', echo=FALSE}
## Iterate through each COSMIC etiology
for (etiology in unique(COSMIC_df$Name)) {
  ## Define signature to use
  sigs <- COSMIC_df[Name == etiology]$Signature
  
  ## Resolve mutation signatures from VCF data
  mut_sig <- resolveMutSignatures(mutCountData = allVCF.counts, signFreqData = cosmix[sigs])
  
  ## Plot COSMIC signatures
  new.p <- msigPlot(x = mut_sig$results$count.result)
  my.feats <- unique(new.p$data$feature)
  new.p +
  ggtitle(label = "COSMIC signatures") +
  scale_fill_manual(name = paste0(etiology, "\nSignatures"),
                    values = grDevices::heat.colors(length(my.feats)))
  
  cat("###", paste0("COSMIC Etiology: ", etiology), " \n")
  print(new.p)
  cat(' \n\n')
}
```

## End